NLTK: The Natural Language Toolkit

### 2002-03-28 technical/acl2002/submission.tex
- Minor revisions - Merged "primary design criteria" and "secondary design criteria"   into a single "requirements" subsection
svn/trunk@530

% To do:
%   - Write up evaluation
%   - Write up other approaches
%   - Write up conclusion
%   - General clean-up
%   - ispell
%   - Should URLs include http:// or not (we should be consistant)?

@@

Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is a difficult
task because different comptational linguistics domains require a
variety of different data structures and functionality; and because a
diverse range of topics may need to be included in the syllabus.

@@

% [EL] ``difficulties'' doesn't seem like the right word here?
Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces, making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

% [EL] akward
In this paper we describe a new approach to the above challenges, a
streamlined and flexible way of organizing the practical component of
an introductory computational linguistics course.  We describe NLTK,
the Natural Language Toolkit, which we have developed in conjunction
with a course we have taught at the University of Pennsylvania.

@@

All materials discussed here are available under an open source
license from \url{nltk.sf.net}.  NLTK runs on most platforms,
including Windows, OS X, Linux, and UNIX.

@@

The most basic step in setting up a practical component is choosing a
suitable programming language.  A number of desiderata influenced our
choice.  First, the language must have a shallow learning curve, so
that novice programmers get immediate rewards for their efforts.
Second, the language must support rapid prototyping and a short
develop/test cycle; an obligatory compilation step is a serious
detraction.  Third, the code should be self-documenting, with a
transparent syntax and semantics.  Fourth, it should be easy to write
structured programs, ideally object-oriented but without the burden
associated with languages like C++.  Finally, the language must have
an easy-to-use graphics library to support the development of simple
graphical user interfaces.

% I moved www.python.org to the \cite.
In surveying the available languages, we believe that Python offers an
especially good fit to the above requirements.  Python is an
object-oriented scripting language developed by Guido van Rossum
and available on all platforms \cite{python}.  Python offers a
shallow learning curve; it was designed to be easily learnt by
children \cite{rossum99}.  As an interpreted language, Python is
suitable for rapid prototyping.  Python code is exceptionally
readable, and it has been praised as ``executable pseudocode.''
% Couldn't find a REF for ``executable pseudocode'' -- closest I got
% was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the ``Tk'' graphics toolkit
\cite{tkinter}, and writing graphical interfaces is straightforward.

@@

Several criteria were considered in the design and implementation of
the toolkit.  The design criteria are divided into primary and
secondary criteria, and listed in the order of their importance.  It
was also important to decide what goals the toolkit would \emph{not}
attempt to accomplish; we therefore include an explicit set of
non-requirements, which the toolkit is not expected to satisfy.

\subsection{Primary Design Criteria}

@@

\paragraph{\textit{Extensibility.}} The toolkit should easily
accommodate new components, whether those components replicate or
extend the toolkit's existing functionality.  Thus, the toolkit's
design should be modular, with simple and well-defined interfaces
between modules.  The toolkit should also be structured in such a way
that it is obvious where new extensions would fit into the toolkit's
infrastructure.

\paragraph{\textit{Documentation.}} The toolkit, its data structures,
and its implementation all need to be carefully and thoroughly
documented.  Three documentation types are necessary: tutorials, to
teach students how to use the toolkit to perform specific tasks;
technical reports, to explain the toolkit's design and implementation;
and reference documentation, to describe every module, interface,
class and method.  All nomenclature must be carefully chosen and
consistently used.

\subsection{Secondary Design Criteria}

@@

\paragraph{\textit{Modularity.}} The interaction between different
components of the toolkit should be kept to a minimum.  In particular,
it should be possible to complete individual projects using small
parts of the toolkit, without worrying about how they interact with
the rest of the toolkit.  This allows students to learn how to use the
toolkit incrementally throughout a course.  Modularity also makes it
easier to change and extend the toolkit.

@@

% ===================== Modules =====================
\section{Modules}
\label{sec:modules}
% What order should these be in?  Put more impressive stuff at the
% beginning and end?

The toolkit is implemented as a collection of independant
\emph{modules}, each of which defines a specific data structure or
task.

A set of core modules defines basic data types and processing systems
that are used througout the toolkit.  The \texttt{token} module
defines basic classes for processing individual elements of text, such
as words or sentences.  The \texttt{tree} module provides data
structures for representing tree structures over text, such as syntax
trees and morphological trees.  The \texttt{probability} module
defines classes that encode frequency distributions and probability
distributions, including a variety of statistical smoothing
techniques.

@@

The \texttt{parser} module defines a high-level interface for
producing trees that represent the structures of texts.  The
\texttt{chunkparser} module defines a subinterface for parsers that
identify non-overlapping linguistic groups (such as noun phrases) in
unrestricted text.

@@

The \texttt{tagger} module defines a standard interface for augmenting
each token of a text with supplementary information, such as its part
of speech or its WordNet synset tag; and provides several different
implementations for that interface.

@@

% jb: this is broken up oddly.
A set of visualization modules define graphical tools for viewing and
editing data structures, and for experimenting with NLP tasks.  The
\texttt{draw.tree} module provides a simple graphical interface for
displaying tree structures; and the
\texttt{draw.tree\_edit} module provides an interface for building and
modifying tree structures.  The \texttt{draw.plot\_graph} module can be
used to graph mathematical functions.  The \texttt{draw.fsa} module
provides a graphical tool for displaying and simulating finite state
automata.  The \texttt{draw.chart} module provides an interactive
graphical tool for experimenting with chart parsers.

@@

The \texttt{classifier} module defines a standard interface for
classifying texts into categories.  This interface is currently
implemented by two modules: the \texttt{classifier.naivebayes} module
defines a text classifier based on the Naive Bayes assumption; and the
\texttt{classifier.maxent} module defines the maximum entropy model
for text classification, and implements two algorithms for training
the model: Generalized Iterative Scaling and Improved Iterative
Scaling.

@@

\paragraph{\textit{Tutorials}} teach students how to use the toolkit,
in the context of performing specific tasks.  Each tutorial focuses on
a single domain, such as tagging, probablistic systems, or text
classification.  The tutorials include a high-level discussion that
explains and motivates the domain, followed by a detailed
walk-through that uses examples to show how NLTK can be used to
perform specific tasks.

@@

\paragraph{\textit{Technical Reports}} explain and justify the
toolkit's design and implementation.  They are used by the developers
of the toolkit to guide and document the toolkit's construction.
Students can also consult these reports if they would like further
information about how the toolkit is designed and why it is designed
that way.

@@

As an example of a moderately difficult assignment, we asked students
to construct a chunk parser that correctly identifies base noun phrase
chunks in a given text, by defining a cascade of transformational
chunking rules.  The NLTK \texttt{rechunkparser} module provides a
variety of rule types, which the students can instantiate to construct
complete rules.  For example, \texttt{ChunkRule('<NN.*>')} builds
chunks from sequences of consecutive nouns; \texttt{ChinkRule('<VB.>')}
excises verbs from existing chunks; \texttt{SplitRule('<NN>', '<DT>')}
splits any existing chunk that contains a singular noun followed by
determiner into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')}
combines two adjacent chunks where the first chunk ends and the second
chunk starts with adjectives.

@@

NLTK provides graphical tools that can be used to give in-class
demonstrations that help to explain basic NLP concepts and algorithms.
These interactive tools can be used to display relevant data
structures, and to show the step-by-step execution of algorithms.
Both data structures and control flow can be easily modified during
the demonstration, in response questions from the class.

@@

% Do we want to mention that there were 40h of lecture time?
% Should we be more explicit about who took the class?
We used NLTK as a basis for the assignments and student projects in
CIS-530, an introductory computational linguistics class taught at the
University of Pennsylvania.  CIS-530 is a graduate level class,
although some advanced undergraduates were also enrolled.  Most
students had either a computer science background or a linguistics
backgroud.  Students were required to complete five assignments, two
exams, and a final project.  All class materials are available from
the corse website \url{http://www.cis.upenn.edu/~cis530/}.

% (Leaving this out for now\ldots)
%The Language: Python was new to 95\% of students.  Some were initially
%bothered about its block structuring, which is sensitive to white
%space.  All quickly learnt the language, and reported that any initial
%reservations were unfounded.  Many students said that they enjoyed
%writing Python code.

% Where does this go??
%At the beginning of the semester, we distributed 2 CD-ROMs to the
%students, containing NLTK and all of the corpera that we would use for
%the class.

@@

We did encounter a few difficulties during the semester.  One problem
is finding large, clean, open-source corpera that the students can use
for their assignments.  This became especially apparent when students
started planning their final projects, since many projects required
different types of corpera.  Another issue was the fact that we were
actively developing NLTK during the semester; some modules were only
completed one or two weeks before we used them.  As a result, students
who worked at home needed to download new versions of the toolkit
several times throughout the semester.  Luckily, Python has extensive
support for installation scripts, which made these upgrades much
simpler.  The students encountered a couple of bugs in the toolkit,
but none were serious, and all were quickly corrected.

@@

%Overview of the paper:
% [EL] Pick better 3 adjectives :)
NLTK provides a simple, extensible, uniform framework for student
assignments and class demonstrations.  It is well documented, easy to
learn, and simple to use.  We hope that NLTK will allow computational
linguistics classes to include more hands-on experience building NLP
components and systems.

% [EL] No room for this for now\ldots
%NOTES: Many CL textbooks provide good exercises at the end of each
%chapter.  This isn't enough for many situations.  E.g. students spend
%lots of time writing interface code (e.g. to process a corpus).
%E.g. teachers spend lots of time creating do-able assignments...

%Future work: 
We plan to continue extending the breadth of materials covered by the
toolkit.  We are currently working on NLTK modules for Hidden Markov
Models, probablistic parsing, and tree adjoining grammars.  We also
plan to increase the number of algorithms implemented by some existing
modules, such as the text classification module.  Finally, we are
putting together a standard collection of large, clean, open-source
corpera, including corpera appropriate for every module defined by the
toolkit.

>>>>

% To do:
%   - Write up evaluation
%   - Write up other approaches
%   - Write up conclusion
%   - General clean-up
%   - ispell
%   - Should URLs include http:// or not (we should be consistant)?
% We assume that the user is familiar with the acronym ``NLP;'' is
% that ok?

@@

% [EL] ``functionality'' isn't a great word, here.
Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is a difficult
task because different comptational linguistics domains require a
variety of different data structures and functionality; and because a
diverse range of topics may need to be included in the syllabus.

@@

% [EL] ``difficulties'' doesn't seem like the right word here?
% Maybe ``challenging requirements''?
Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces, making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

In this paper we introduce a new approach to the above challenges, a
streamlined and flexible way of organizing the practical component of
an introductory computational linguistics course.  We describe NLTK,
the Natural Language Toolkit, which we have developed in conjunction
with a course we have taught at the University of Pennsylvania.

@@

% [EL] This was too ambiguous for me (e.g., we discuss Python)
%All materials discussed here are available under an open source
%license from \url{nltk.sf.net}.  NLTK runs on most platforms,
%including Windows, OS X, Linux, and UNIX.
The Natural Language Toolkit is available under an open source license
from \url{http://nltk.sf.net}.  NLTK runs on most platforms, including
Windows, OS X, Linux, and UNIX.

@@

% [EL] is ``desiderata'' an acceptable word here?
The most basic step in setting up a practical component is choosing a
suitable programming language.  A number of desiderata influenced our
choice.  First, the language must have a shallow learning curve, so
that novice programmers get immediate rewards for their efforts.
Second, the language must support rapid prototyping and a short
develop/test cycle; an obligatory compilation step is a serious
detraction.  Third, the code should be self-documenting, with a
transparent syntax and semantics.  Fourth, it should be easy to write
structured programs, ideally object-oriented but without the burden
associated with languages like C++.  Finally, the language must have
an easy-to-use graphics library to support the development of
graphical user interfaces.

% [EL] ``punitive'' like a strange word.
% [EL] I couldn't find a REF for ``executable pseudocode'' -- closest
% I got was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
In surveying the available languages, we believe that Python offers an
especially good fit to the above requirements.  Python is an
object-oriented scripting language developed by Guido van Rossum
and available on all platforms \cite{python}.  Python offers a
shallow learning curve; it was designed to be easily learnt by
children \cite{rossum99}.  As an interpreted language, Python is
suitable for rapid prototyping.  Python code is exceptionally
readable, and it has been praised as ``executable pseudocode.''
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the ``Tk'' graphics toolkit
\cite{tkinter}, and writing graphical interfaces is straightforward.

@@

% [EL] Is this ambiguous between descending/ascending?
Several criteria were considered in the design and implementation of
the toolkit.  These design criteria are listed in the order of their
importance.  It was also important to decide what goals the toolkit
would \emph{not} attempt to accomplish; we therefore include an
explicit set of non-requirements, which the toolkit is not expected to
satisfy.

\subsection{Requirements}

@@

\paragraph{\textit{Extensibility.}} The toolkit should easily
accommodate new components, whether those components replicate or
extend the toolkit's existing functionality.  The toolkit should be
structured in such a way that it is obvious where new extensions would
fit into the toolkit's infrastructure.

% [EL] I commented out the list of documentation types, since we have
% a section that talks about the documentation.
\paragraph{\textit{Documentation.}} The toolkit, its data structures,
and its implementation all need to be carefully and thoroughly
documented.  
%Three documentation types are necessary: tutorials, to
%teach students how to use the toolkit to perform specific tasks;
%technical reports, to explain the toolkit's design and implementation;
%and reference documentation, to describe every module, interface,
%class and method.  
All nomenclature must be carefully chosen and consistently used.

@@

\paragraph{\textit{Modularity.}} The interaction between different
components of the toolkit should be kept to a minimum, using simple,
well-defined interfaces.  In particular, it should be possible to
complete individual projects using small parts of the toolkit, without
worrying about how they interact with the rest of the toolkit.  This
allows students to learn how to use the toolkit incrementally
throughout a course.  Modularity also makes it easier to change and
extend the toolkit.

@@

% ===================== Modules =====================
\section{Modules}
\label{sec:modules}
% What order should the subsections be in?  Put more impressive stuff
% at the beginning and end?

% [EL] is ``defines'' the right word?
The toolkit is implemented as a collection of independant
\emph{modules}, each of which defines a specific data structure or
task.

A set of core modules defines basic data types and processing systems
that are used througout the toolkit.  The \texttt{token} module
provides basic classes for processing individual elements of text,
such as words or sentences.  The \texttt{tree} module defines data
structures for representing tree structures over text, such as syntax
trees and morphological trees.  The \texttt{probability} module
implements classes that encode frequency distributions and probability
distributions, including a variety of statistical smoothing
techniques.

@@

The \texttt{parser} module defines a high-level interface for
producing trees that represent the structures of texts.  The
\texttt{chunkparser} module defines a subinterface for parsers that
identify non-overlapping linguistic groups (such as base noun phrases)
in unrestricted text.

@@

The \texttt{tagger} module defines a standard interface for augmenting
each token of a text with supplementary information, such as its part
of speech or its WordNet synset tag; and provides several different
implementations for this interface.

@@

% jb: this is broken up oddly.

Visualization modules define graphical interfaces for viewing and
manipulating data structures, and graphical tools for experimenting
with NLP tasks.  The \texttt{draw.tree} module provides a simple
graphical interface for displaying tree structures; and the
\texttt{draw.tree\_edit} module provides an interface for building and
modifying tree structures.  The \texttt{draw.plot\_graph} module can be
used to graph mathematical functions.  The \texttt{draw.fsa} module
provides a graphical tool for displaying and simulating finite state
automata.  The \texttt{draw.chart} module provides an interactive
graphical tool for experimenting with chart parsers.

@@

The \texttt{classifier} module defines a standard interface for
classifying texts into categories.  This interface is currently
implemented by two modules.  The \texttt{classifier.naivebayes} module
defines a text classifier based on the Naive Bayes assumption. The
\texttt{classifier.maxent} module defines the maximum entropy model
for text classification, and implements two algorithms for training
the model: Generalized Iterative Scaling and Improved Iterative
Scaling.

@@

% [EL] Provide some indication of how much tutorial material is
% available?  (Currently, 8 tutorials, mean=16pp, range=6pp-52pp.)
\paragraph{\textit{Tutorials}} teach students how to use the toolkit,
in the context of performing specific tasks.  Each tutorial focuses on
a single domain, such as tagging, probablistic systems, or text
classification.  The tutorials include a high-level discussion that
explains and motivates the domain, followed by a detailed
walk-through that uses examples to show how NLTK can be used to
perform specific tasks.

@@

\paragraph{\textit{Technical Reports}} explain and justify the
toolkit's design and implementation.  They are used by the developers
of the toolkit to guide and document the toolkit's construction.
Students can also consult these reports if they would like further
information about how the toolkit is designed, and why it is designed
that way.

@@

As an example of a moderately difficult assignment, we asked students
to construct a chunk parser that correctly identifies base noun phrase
chunks in a given text, by defining a cascade of transformational
chunking rules.  The NLTK \texttt{rechunkparser} module provides a
variety of regular-expression based rule types, which the students can
instantiate to construct complete rules.  For example,
\texttt{ChunkRule('<NN.*>')} builds chunks from sequences of
consecutive nouns; \texttt{ChinkRule('<VB.>')} excises verbs from
existing chunks; \texttt{SplitRule('<NN>', '<DT>')} splits any
existing chunk that contains a singular noun followed by determiner
into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')} combines two
adjacent chunks where the first chunk ends and the second chunk starts
with adjectives.

@@

NLTK provides graphical tools that can be used in class demonstrations
to help explain basic NLP concepts and algorithms.  These interactive
tools can be used to display relevant data structures and to show the
step-by-step execution of algorithms.  Both data structures and
control flow can be easily modified during the demonstration, in
response questions from the class.

@@

% Do we want to mention that there were 40h of lecture time?
% Should we be more explicit about who took the class?
We used NLTK as a basis for the assignments and student projects in
CIS-530, an introductory computational linguistics class taught at the
University of Pennsylvania.  CIS-530 is a graduate level class,
although some advanced undergraduates were also enrolled.  Most
students had a background in either computer science or linguistics
(but not both).  Students were required to complete five assignments,
two exams, and a final project.  All class materials are available
from the corse website \url{http://www.cis.upenn.edu/~cis530/}.

% (Leaving this out for now..)
%The Language: Python was new to 95\% of students.  Some were initially
%bothered about its block structuring, which is sensitive to white
%space.  All quickly learnt the language, and reported that any initial
%reservations were unfounded.  Many students said that they enjoyed
%writing Python code.

% Where does this go??  Leaving it out for now..
%At the beginning of the semester, we distributed 2 CD-ROMs to the
%students, containing NLTK and all of the corpera that we would use for
%the class.

@@

We did encounter a few difficulties during the semester.  One problem
was finding large clean corpera that the students could use for their
assignments.  Several of the students needed assistance finding
suitable corpera for their final projects.  Another issue was the fact
that we were actively developing NLTK during the semester; some
modules were only completed one or two weeks before the students used
them.  As a result, students who worked at home needed to download new
versions of the toolkit several times throughout the semester.
Luckily, Python has extensive support for installation scripts, which
made these upgrades much simpler.  The students encountered a couple
of bugs in the toolkit, but none were serious, and all were quickly
corrected.

@@

%Overview of the paper:
% [EL] Pick better 3 adjectives :)
NLTK provides a simple, extensible, uniform framework for student
assignments and class demonstrations.  It is well documented, easy to
learn, and simple to use.  We hope that NLTK will allow computational
linguistics classes to include more hands-on experience with using and
building NLP components and systems.

% [EL] No room for this for now...
%NOTES: Many CL textbooks provide good exercises at the end of each
%chapter.  This isn't enough for many situations.  E.g. students spend
%lots of time writing interface code (e.g. to process a corpus).
%E.g. teachers spend lots of time creating do-able assignments...

%Future work: 
We plan to continue extending the breadth of materials covered by the
toolkit.  We are currently working on NLTK modules for Hidden Markov
Models, probablistic parsing, and tree adjoining grammars.  We also
plan to increase the number of algorithms implemented by some existing
modules, such as the text classification module.

Finding suitable corpera is a prerequisite for many studnet
assignments and projects.  We are therefore putting together a
standard collection of large, clean, open-source corpera, containing
corpera appropriate for every module defined by the toolkit.
